---
title: "Principal Component Analysis (PCA)"
subtitle: "Tutorial"
date: "`r Sys.Date()`"
author: "Orlando Sabogal-Cardona"
theme:
  light: journal
  dark: darkly
format:
  html:
    toc: true
    toc_levels: 3
editor: visual
---

#Libraries and data

```{r}
library(tidyverse)
library(magrittr)

library(factoextra)
library(psych)
```


# First example

## Iris data

```{r}
names(iris)
```
```{r}
MyData <- iris %>% select(-Species)
```


## prcomp 

### prcomp (scale = FALSE)

```{r}
MyFirst_PCA <- prcomp(MyData)
```

```{r}
MyFirst_PCA
```

```{r}
str(MyFirst_PCA)
```

**$rotation** is the matrix of the principal component loadings (eigenvectors). Each column of **rotation** contains coefficients for one principal component. 

```{r}
MyFirst_PCA$rotation
```

The eigenvalues are a bit tricky. In **$sdev** you will find the standard deviation of each principal component. It provides a measure of the amount of variation captured by each component. Remember that the igenvalues are associated with the variance of the components. Therefore, you only need to *^2* 


```{r}
MyFirst_PCA$sdev
```

```{r}
MyFirst_PCA$sdev^2
```

```{r}
MyFirst_PCA$sdev^2/sum(MyFirst_PCA$sdev^2)
```

```{r}
summary(MyFirst_PCA)
```   

The **"scores"** of the components:

```{r}
MyFirst_PCA$x %>% head()
```


### prcomp (scale = TRUE)

```{r}
MyFirst_PCA <- prcomp(MyData, scale = TRUE)
```

```{r}
MyFirst_PCA
```

```{r}
MyFirst_PCA$rotation
```

```{r}
summary(MyFirst_PCA)
```


## princomp


```{r}
PCA_Results <- princomp(MyData, cor = TRUE, scores = TRUE)
```

```{r}
PCA_Results
```

```{r}
summary(PCA_Results)
```

```{r}
str(PCA_Results)
```
```{r}
PCA_Results$loadings
```

```{r}
PCA_Results$scores %>% head()
```

## prcomp vs princomp


The **prcomp()** function utilizes something calles Spectral Value Decomposition SVD. 
The **pincomp()** function utilizes Spectra Decomposition. 


## factorextra


```{r}
MyFirst_PCA <- prcomp(MyData, scale = TRUE)
```

```{r}
fviz_eig(MyFirst_PCA)
```


```{r}
fviz_pca_ind(MyFirst_PCA)
```

```{r}
fviz_pca_var(MyFirst_PCA)
```

```{r}
fviz_pca_biplot(MyFirst_PCA,
                col.var = "red", 
                col.ind = "green")
```


## With psych


## By hand



# Second example









